\documentclass[10pt,letterpaper]{article}
\usepackage[letterpaper,margin=0.75in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[russian,english]{babel}
\usepackage{mdwlist}
\usepackage[T2A]{fontenc}
\usepackage{textcomp}
\usepackage{tgpagella}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{tikz-3dplot}
\usepackage{caption}
\usepackage{subfig}
\usepackage{indentfirst}
\usepackage{listings}
\setlength{\tabcolsep}{2em}
\setlength{\parindent}{4em}
\setlength{\parskip}{1em}

\numberwithin{equation}{section}
\numberwithin{figure}{section}
\numberwithin{table}{section}
\renewcommand{\thesubfigure}{\thefigure.\arabic{subfigure}}
\captionsetup[subfigure]{labelformat=simple,labelsep=colon,listofformat=subsimple}

\title{Coordinates conversion in acoustic positioning}
\author{Oleksandr Novychenko}

\begin{document}

\maketitle

\section{General information}
During an acoustic positioning by ultra-short baseline transceiver a direction to a
target is measured as a round-trip time. When sound speed is known a distance between
USBL transceiver and transponder can be calculated. In combination with measured direction
it allows to calculate target coordinates in USBL-frame (or USBL-Frame) $\vec{X}_{ut}$
in \texttt{USBLLONG}.

In some scenarios USBL transceiver can't measure distance to target. In that case bearing
$\theta_{ut}$ and elevation $\varphi_{ut}$ angles are provided in \texttt{USBLANGLES}.
Covertion to USBL-Frame when distance $d$ is known is:
\begin{equation}
    \vec{X}_{ut} = d \cdot \begin{bmatrix}
        \sin \theta_{ut} \cos \varphi_{ut} \\
        \cos \theta_{ut} \cos \varphi_{ut} \\
        \sin \varphi_{ut}
    \end{bmatrix}
\end{equation}

USBL-Frame corrdinates $\vec{X}_{ut}$ should be converted to Vessel-Frame coordinates
$\vec{X}_vt$ by correction to USBL location $\vec{X}_{vu}$ and rotation $[R_{vu}]$:
\begin{equation}
    \vec{X}_{vt} = \vec{X}_{vu} + [R_{vu}] \cdot \vec{X}_{ut}.
    \label{equ:lf2vf}
\end{equation}

Vessel-Frame coordinates of target $\vec{X}_{vt}$ could be motion-compensated by using
information about attitude and heading of a vessel $[R_{vn}]$. Let $\varphi_{vn}$ (roll), $\theta_{vn}$
(pitch) and $\psi_{vn}$ (yaw) angles are provided by vessels AHRS. As result Topocentric
coordinates (North-East-Down) with Vessel-Frame origin are computed:
\begin{equation}
    \vec{N}_{vt} = [R_{vn}] \cdot \vec{X}_{vt}
    \label{equ:vf2ned}
\end{equation}

When Geodetic coordinates of Vessel-Frame origin $\vec{G}_{v}$ and Topocentric coordinates
of target $\vec{N}_{vt}$ are known, Geodetic coordinates of target $\vec{G}_{t}$ can be
computed.


\section{Definitions}

\subsection{Coordinate systems}
Vessel-Frame is cartesian coordinate system with origin somewere on a vessel. Usually on
stern. $X_{v}$-axis is directed forward, $Y_{v}$-axis directed astarboard and $Z_{v}$-axis
is directed down by right-hand rule.

USBL-Frame coordinate system has origin in the middle of USBL transceiver. $Y_{u}$-axis
is directed to white mark on USBL-head (forward), $Z_{u}$-axis is directed to subconn port
direction (up) and $X_{u}$ by right-hand rule directs right.

\begin{figure}
    \centering
    \tdplotsetmaincoords{70}{30}
    \begin{tikzpicture}[scale=5,tdplot_main_coords]
        \coordinate (O) at (0,0,0);
        \coordinate (U) at (0.1,0.03,-0.15);

        \draw[] (0,0,0) node[anchor=south east]{$O$};
        \draw[thick,->] (0,0,0) -- (0,1,0) node[anchor=south west]{$X_{v}$};
        \draw[thick,->] (0,0,0) -- (1,0,0) node[anchor=north west]{$Y_{v}$};
        \draw[thick,->] (0,0,0) -- (0,0,-0.75) node[anchor=north]{$Z_{v}$};

        \draw[thin,dashed,->] (O) -- (U) node[anchor=north]{$\vec{X}_{vu}$};

        \tdplotsetrotatedcoordsorigin{(U)};
        \draw[dotted,tdplot_rotated_coords,->] (0,0,0) -- (0.5,0,0);
        \draw[dotted,tdplot_rotated_coords,->] (0,0,0) -- (0,0.5,0);
        \draw[dotted,tdplot_rotated_coords,->] (0,0,0) -- (0,0,0.5);

        \tdplotsetrotatedcoords{-15}{5}{0};
        \draw[thick,tdplot_rotated_coords,->] (0,0,0) -- (0.5,0,0) node[anchor=west]{$X_{u}$};
        \draw[thick,tdplot_rotated_coords,->] (0,0,0) -- (0,0.5,0) node[anchor=south]{$Y_{u}$};
        \draw[thick,tdplot_rotated_coords,->] (0,0,0) -- (0,0,0.5) node[anchor=west]{$Z_{u}$};
    \end{tikzpicture}
    \caption{USBL-Frame located in Vessel-Frame}
    \label{pic:devcrf}
\end{figure}


\subsection{Rotation}
Rotation of coordinates around speceific axis on specific angle could be done by using
Quaternions or Direction-Cosine Matrices (or Rotation Matrices).

Quaternions allows to apply rotation around arbitary axes. Positive direction of rotation
is a clockwise rotation when watching towards the axis. DCM allows the same, but its
more intuitive to describe vector operations with it.

Multiplying rotation Matrix to a column vector produces rotated column vector.

DCM can be built from quaternion in case of arbitary axis $\vec{r}$ and angle $\Omega$ or
as a combination of rotations around $X$, $Y$ and $Z$ axes (in $X-Y-Z$ order).

Angles $\varphi$, $\theta$ and $\psi$ are also known as Euler angles. Rotation matrix
$[R(\varphi, \theta, \psi)] = [R_{z}(\psi)] \cdot [R_{y}(\theta)] \cdot [R_{x}(\varphi)]$.

Applying rotation $[R_{ab}]$ and shift $\vec{X}_{ab}$ to point $\vec{X}_{a}$:
\begin{equation}
    \begin{bmatrix}
        x_{b} \\ y_{b} \\ z_{b}
    \end{bmatrix} = \begin{bmatrix}
        x_{ab} \\ y_{ab} \\ z_{ab}
    \end{bmatrix} + \begin{bmatrix}
         \cos \psi_{ab}   & -\sin \psi_{ab}    & 0                  \\
         \sin \psi_{ab}   &  \cos \psi_{ab}    & 0                  \\
        0                 & 0                  & 1
    \end{bmatrix} \cdot \begin{bmatrix}
         \cos \theta_{ab} & 0                  &  \sin \theta_{ab}  \\
        0                 & 1                  & 0                  \\
        -\sin \theta_{ab} & 0                  &  \cos \theta_{ab}
    \end{bmatrix} \cdot \begin{bmatrix}
        1                 & 0                  & 0                  \\
        0                 &  \cos \varphi_{ab} & -\sin \varphi_{ab} \\
        0                 &  \sin \varphi_{ab} &  \cos \varphi_{ab}
    \end{bmatrix} \cdot \begin{bmatrix}
        x_{a} \\ y_{a} \\ z_{a}
    \end{bmatrix}.
    \label{equ:rot_xyz}
\end{equation}

\begin{figure}
    \centering
    \tdplotsetmaincoords{60}{30}
    \begin{tikzpicture}[scale=5,tdplot_main_coords]
        \draw[] (0,0,0)  node[anchor=east]{$O$};
        \draw[thick,->] (0,0,0) -- (1,0,0) node[anchor=north west]{$X$};
        \draw[thick,->] (0,0,0) -- (0,1,0) node[anchor=south west]{$Y$};
        \draw[thick,->] (0,0,0) -- (0,0,1) node[anchor=south]{$Z$};

        \tdplotsetrotatedcoords{0}{90}{0};
        \tdplotdrawarc[thin,tdplot_rotated_coords,->]{(0,0,0.66)}{0.1}{0}{270}{anchor=south}{$\varphi$};

        \tdplotsetrotatedcoords{90}{90}{60};
        \tdplotdrawarc[thin,tdplot_rotated_coords,->]{(0,0,0.66)}{0.1}{0}{270}{anchor=south}{$\theta$};

        \tdplotsetrotatedcoords{90}{0}{90};
        \tdplotdrawarc[thin,tdplot_rotated_coords,->]{(0,0,0.66)}{0.1}{0}{270}{anchor=north west}{$\psi$};
    \end{tikzpicture}
    \caption{Euler angles}
    \label{pic:euler}
\end{figure}

Single forlmula to computation rotation matrix (\ref{equ:rot_xyz}) from Euler angles:
\begin{equation}
    [R(\varphi, \theta, \psi)] = \begin{bmatrix}
        \cos \theta \cos \psi &
        \sin \varphi \sin \theta \cos \psi - \cos \varphi \sin \psi &
        \cos \varphi \sin \theta \cos \psi + \sin \varphi \cos \psi \\
        \cos \theta \sin \psi &
        \sin \varphi \sin \theta \sin \psi + \cos \varphi \cos \psi &
        \cos \varphi \sin \theta \sin \psi - \sin \varphi \cos \psi \\
        -\sin \theta &
        \sin \varphi \cos \theta &
        \cos \varphi \cos \theta
    \end{bmatrix}.
    \label{equ:rpy2dcm}
\end{equation}

Angles can be extracted from DCM by:
\begin{equation}
    \begin{split}
        \varphi & = \arctg R_{32} / R_{33}, \\
        \theta  & = -\arcsin R_{31}, \\
        \psi    & = \arctg R_{21} / R_{11}.
    \end{split}
    \label{equ:dcm2rpy}
\end{equation}

DCMs has a property: $[R]^{-1} = [R]^{T}$.

From physical point of view, $\varphi$, $\theta$ and $\psi$ angles are also called as
Roll, Pitch and Yaw. Usually AHRS sensors provides angles for NED-frame: Roll --- port up,
Pitch --- bow up, Yaw --- heading for Vessel-Frame specified earlier.


\section{Coordinates convertion}


\subsection{Topocentric coordinates}

As mentioned in (\ref{equ:lf2vf}) an offset and rotation of USBL relaive to
Vessel-frame is required.

It is possible that AHRS sensor located with some bias $[R_{va}]$. In this case
Euler angles (or rotation matrix $[R_{an}]$) convertion is needed.

As mentionoed in (\ref{equ:vf2ned}), coordinates in AHRS-Frame can be converted
to $NED$ by multiplication to $[R_{vn}]$. To convert coordinates from Vessel-Frame
to AHRS Frame a multiplication to $[R_{va}]^{-1}$ is needed.

Inverted DCM $[R_{va}]^{-1}$ has an easy explanation. Since $[R_{vu}]$ (USBL-Frame
rotation) and $[R_{va}]$ (AHRS-Frame rotation) are both specified in Vessel-Frame,
first is used to convert $USBL \rightarrow VF$, second for $AHRS \rightarrow VF$.
Since transformation $VF \rightarrow AHRS$ is needed, rotation matrix is inverted.
\begin{equation}
    [R_{vn}] = [R_{an}] \cdot [R_{va}]^{-1}
\end{equation}


\subsection{Geographic coordinates}

When GNSS is used, geodetic coordinates $\vec{G}_{g}$ of antenna (which is
located at $\vec{X}_{vg}$ in Vessel-Frame) are measured:
\begin{equation}
    \vec{G} = [ \varphi, \lambda, h ]^{T}
\end{equation}

Geogetic Latitude $\varphi$ and Longitude $\lambda$ will be interpreted as in
radians and in WGS-84 datum.

All geodetic computations are done in Earth-centered Earth-fixed (ECEF) coordinates.
Origin of ECEF is located in center os mass of Earth. X-axis is directed to Greenwich
meridian ($\varphi = 0$, $\lambda = 0$), Y-axis is directed to North-pole ($\varphi = \pi$)
and meters are used as units.

Convertion between Gerodetic and ECEF coordinates are done by EPSG:9602.
Let ellipsoid curvature on specified Latitude is:
\begin{equation}
    \nu = \frac{a}{\sqrt{1 - e^2 \sin^2 \varphi}}
\end{equation}
where $a$ -- major semiaxis and $e$ -- excentricity of WGS-84 ellipsoid. These constants
can be found in table \ref{tab:ellps}.

Geodetic coordinates convertion to ECEF is:
\begin{equation}
    \begin{split}
        X & = (\nu + h) \cdot cos \varphi \cdot cos \lambda \\
        Y & = (\nu + h) \cdot cos \varphi \cdot sin \lambda \\
        Z & = ((b / a)^2 \cdot \nu + h) \cdot sin \varphi
    \end{split}
    \label{equ:geod2ecef}
\end{equation}

ECEF coordinates convertion to Geodetic is a tricky process, but EPSG:9602 proposes
iteration-less formulas:
\begin{equation}
    \begin{split}
        \varphi & = \arctan \frac{Z + \varepsilon^{'} \cdot b \cdot \sin^3 q}{p - e^2 \cdot a \cdot \cos^3 q} \\
        \lambda & = \arctan \frac{Y}{X}
    \end{split}
    \label{equ:ecef2geod}
\end{equation}
where $\varepsilon^{'} = e^2 / (1 - e^2)$ -- second excentricity, $p = \sqrt{X^2 + Y^2}$ and
$q = \arctan \frac{Z \cdot a}{p \cdot b}$. Note, that \texttt{atan2()} should be used for
$\lambda$ computation.

\begin{table}[h]
    \centering
    \begin{tabular}{  c | c | c | c | c  }
        \hline
        Ellipsoid & $a=R_{maj}$ & $b=R_{min} $ & $f=(a-b)/a$ & $e=\sqrt{1-(b/a)^{2}}$ \\
        \hline
        WGS-84 & 6378137 & 6356752.3142 & 1/298.257223563 & 0.0818191909289069 \\
        PZ-90  & 6378136 & 6356751.3617 & 1/298.257839303 & 0.0818191066156667 \\
        \hline
    \end{tabular}
    \caption{Global ellipsoids paramaters}
    \label{tab:ellps}
\end{table}

Rotation from Topocentric ($ENU$) frame to ECEF at specified coordinates $\vec{G}_{t}$
is done by EPSG:9836:
\begin{equation}
    [R_{te}(\varphi, \lambda)] = \begin{bmatrix}
        -\sin \lambda & -\sin \varphi \cos \lambda & \cos \varphi \cos \lambda \\
         \cos \lambda & -\sin \varphi \sin \lambda & \cos \varphi \sin \lambda \\
         0            &  \cos \varphi              & \sin \varphi
    \end{bmatrix}
    \label{equ:geod2dcm}
\end{equation}

To find geodetic coordinates of the point with $ENU$-coordinates relative to
$\vec{G}_0$, it needed to be converted to ECEF $[ X_0, Y_0, Z_0 ]^{T}$ and:
\begin{equation}
    \begin{bmatrix}
        X \\ Y \\ Z
    \end{bmatrix} = [R_{te}(\varphi_0, \lambda_0)] \cdot \begin{bmatrix}
        E \\ N \\ U
    \end{bmatrix} + \begin{bmatrix}
        X_0 \\ Y_0 \\ Z_0
    \end{bmatrix}
\end{equation}

Let convertion (\ref{equ:geod2ecef}) will be $\vec{E} = toECEF(\vec{G})$,  convertion
(\ref{equ:ecef2geod}) will be $\vec{G} = fromECEF(\vec{E})$ and (\ref{equ:geod2dcm}) will
be $[R_{te}] = [rotECEF(\vec{G})]$. Then:
\begin{equation}
    \vec{G}_{t} = fromECEF(toECEF(\vec{G}_{v}) + rotECEF(\vec{G}_{v}) \cdot [F] \cdot \vec{N}_{vt})
    \label{equ:ned2geod}
\end{equation}
where $\vec{G}_{t}$ -- geodetic target coordinates, $\vec{G}_{v}$ -- geodetic Vessel-Frame
origin cordinates, $\vec{N}_{vt}$ -- target $NED$-coordinates in Vessel-Frame and $[F]$ is
flip-matrix for $NED \rightarrow ENU$ convertion:
\begin{equation}
    [F] = \begin{bmatrix}
        0 & 1 & 0 \\ 1 & 0 & 0 \\ 0 & 0 & -1
    \end{bmatrix}
\end{equation}

\subsection{Final algorithm}

The combination of (\ref{equ:lf2vf}), (\ref{equ:vf2ned}) and (\ref{equ:ned2geod}) with assumption,
that GNSS antenna is located in Vessel-Frame origin is:
\begin{equation}
    \begin{split}
        \vec{X}_{vt} & = \vec{X}_{vu} + [R_{vu}] \cdot [F] \cdot \vec{X}_{ut}, \\
        \vec{N}_{vt} & = [R_{an}] \cdot [R_{va}]^{T} \cdot \vec{X}_{vt}, \\
        \vec{G}_{t}  & = fromECEF(toECEF(\vec{G}_{v}) + rotECEF(\vec{G}_{v}) \cdot [F] \cdot \vec{N}_{vt}),
    \end{split}
\end{equation}
where:
\begin{itemize}
    \item $\vec{X}_{ut}$ -- coordinates of target extracted from \texttt{USBLLONG};
    \item $\vec{X}_{vu}$ -- USBL offset in Vessel-Frame;
    \item $[R_{vu}]$ -- rotation of USBL-Frame in Vessel-Frame;
    \item $[R_{an}]$ -- rotation of AHRS-frame to NED-frame;
    \item $[R_{va}]$ -- rotation AHRS in Vessel-Frame;
    \item $\vec{G}_{v}$ -- geodetic coordinates of Vessel-Frame origin.
\end{itemize}

Note the $[F]$ application to $\vec{X}_{ut}$. In defference from (\ref{equ:lf2vf}),
axes difference (Fig.\ref{pic:devcrf}) is taken into account.

%\lstinputlisting{../lib/get_n.m}
%\lstinputlisting{../lib/geod2ecef.m}
%\lstinputlisting{../lib/ecef2geod.m}

\end{document}
